### 2021-04-15 submission/manuscript.Rmd
Add to results section

**5-day laxative treatment leads to prolonged *C. difficile* colonization in mice.** We compared PEG-treated mice to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile*, with mice typically clearing *C. difficile* within 10 days post-infection [@Tomkovich2019; @Tomkovich2020]. All PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, one group was also treated with clindamycin, and one group was allowed to recover for 10 days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment (Fig. 1B). After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^3^ *C. difficile* 630 spores. All treatments rendered mice susceptible to *C. difficile* colonization (Fig. 1C), however PEG-treated mice remained colonized at a high level through 30 days post-infection. In contrast, the clindamycin-treated mice that cleared *C. difficile* within 10 days post-infection. Surprisingly, mice were still susceptible to *C. difficile* infection after 10-days of recovery from PEG treatment although *C. difficile* was not detectable in most of the group in the initial 5 days post-infection (Fig. 1C). From 9 days post-infection onward, the median *C. difficile* stabilized for the 5-day PEG plus 10-day recovery group of mice and remained high through 30 days post-infection (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible for up to 10 days post PEG treatment.

@@

f_d_top_contrib= (f_d_top_2_contrib[1,1])
f_d_sec_contrib = (f_d_top_2_contrib[2,1])
f_d_top_R2_sum = f_d_top_2_contrib %>% 
  select(R2) %>% sum
f_d_top_p <- f_d_top_2_contrib %>% 
  select(`Pr(>F)`) %>% max # Both p values are the same, will be rounded in text
f_d_top_R2 = f_d_top_2_contrib[1,2]
f_d_sec_R2 = f_d_top_2_contrib[2,2]
```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C. difficile* colonization dynamics between mice treated with laxatives versus clindamycin were due to their differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`) with the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (Data Set S1, Sheet X). The impact of cage effects has been well documented due to copraphagy and is expected.

@@

**Figure 1. 5-day PEG treatment prolongs susceptibility and mice become persistently colonized with *C. difficile*.** A. Setup of the experimental timeline for experiments with 5-day PEG treated mice. Clindamycin was administered at 10 mg/kg by intraperitoneal injection. 15% PEG 3350 was administered in the drinking water for five days. B. Weight change from baseline weight in groups after treatment with PEG and/or clindamycin, followed by *C. difficile* challenge. C. *C. difficile* CFU/gram stool measured over time (N = `r five_d_cfu_n_min`-`r five_d_cfu_n_max` mice per timepoint) via serial dilutions. The black line represents the limit of detection for the first serial dilution. CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths. For B-C, lines represent the median for each treatment group and circles represent samples from individual mice. Asterisks indicate timepoints where the weight change or CFU/g was significantly different between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple timepoints.
\newpage

@@

``` {r figure_4_permnova}
one_day_permnova_results <- read_tsv(here("data/process/1_day_PEG_permanova.tsv"))
one_day_top_2_contrib <- one_day_permnova_results %>% arrange(desc(R2)) %>%
  top_n(3, R2)
one_day_top_contrib_name = (one_day_top_2_contrib[2,1])
one_day_sec_contrib_name = (one_day_top_2_contrib[3,1])
top_R2 = one_day_top_2_contrib[2,6]
sec_R2 = one_day_top_2_contrib[3,6]

>>>>

**5-day laxative treatment leads to prolonged *C. difficile* colonization in mice.** We compared mice treated with the osmotic laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile*, with mice typically clearing *C. difficile* within 10 days post-infection [@Tomkovich2019; @Tomkovich2020]. All PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, one group was also treated with clindamycin, and one group was allowed to recover for 10 days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment and the mice regained most of the weight five days after treatment (Fig. 1B). After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^3^ *C. difficile* 630 spores. All treatments rendered mice susceptible to *C. difficile* colonization (Fig. 1C), however PEG-treated mice remained colonized at a high level through thirty days post-infection. In contrast, the clindamycin-treated mice that cleared *C. difficile* within ten days post-infection. Surprisingly, PEG-treated mice were still susceptible to *C. difficile* infection after 10-days of recovery from treatment although *C. difficile* was not detectable in most of the group in the initial five days post-infection (Fig. 1C). From 9 days post-infection onward, the median *C. difficile* stabilized for the 5-day PEG plus 10-day recovery group of mice and remained high through 30 days post-infection (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible for up to ten days post-treatment.

@@

f_d_top_contrib= (f_d_top_2_contrib[1,1])
f_d_sec_contrib = (f_d_top_2_contrib[2,1])
f_d_top_R2_sum = f_d_top_2_contrib %>% 
  select(R2) %>% sum
f_d_top_p <- f_d_top_2_contrib %>% 
  select(`Pr(>F)`) %>% max # Both p values are the same, will be rounded in text
f_d_top_R2 = f_d_top_2_contrib[1,2]
f_d_sec_R2 = f_d_top_2_contrib[2,2]

#Number of experiments represented in 5-day subset
#Use same code from five_day_PEG_subset() in code/utilities.R to determine number of samples we had C. difficile CFU counts for per day
five_day_mice <- metadata %>% filter(group == "C" & exp_num %in% c("M3","M4", "M5", "M8")| #Only use C mice from these experiments. Allocated groups to figures based on paper outline.
             group == "WM" & exp_num %in% c("M3","M4", "M5", "M8")|
             group == "WMC" & exp_num %in% c("M3","M4")|
             group == "WMR" & exp_num %in% c("M5","M6"))
#Total number of experiments in 5-day subset
five_day_total_exp <- five_day_mice %>% 
  filter(!duplicated(exp_num)) %>% #Remove duplicate experiment numbers
  tally() #Total number of unique experiments

```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C. difficile* colonization dynamics between mice treated with laxatives versus clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`) with the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy, we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). When we examine alpha diversity by looking at Shannon diversity index over time, we see that although both clindamycin and PEG treatments decrease diversity, the diversity index is less in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection.

@@

**Figure 1. 5-day PEG treatment prolongs susceptibility and mice become persistently colonized with *C. difficile*.** A. Setup of the experimental timeline for experiments with 5-day PEG treated mice. Clindamycin was administered at 10 mg/kg by intraperitoneal injection. 15% PEG 3350 was administered in the drinking water for five days. B. Weight change from baseline weight in groups after treatment with PEG and/or clindamycin, followed by *C. difficile* challenge. C. *C. difficile* CFU/gram stool measured over time (N = `r five_d_cfu_n_min`-`r five_d_cfu_n_max` mice per timepoint) via serial dilutions. The black line represents the limit of detection for the first serial dilution. CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths. For B-C, lines represent the median for each treatment group and circles represent samples from individual mice. Asterisks indicate timepoints where the weight change or CFU/g was significantly different between groups by the Kruskal-Wallis test with Benjamini-Hochberg correction for testing multiple timepoints. The data presented are from a total of `r five_day_total_exp` separate experiments. 
\newpage

@@

``` {r figure_4_permanova}
one_day_permnova_results <- read_tsv(here("data/process/1_day_PEG_permanova.tsv"))
one_day_top_2_contrib <- one_day_permnova_results %>% arrange(desc(R2)) %>%
  top_n(3, R2)
one_day_top_contrib_name = (one_day_top_2_contrib[2,1])
one_day_sec_contrib_name = (one_day_top_2_contrib[3,1])
top_R2 = one_day_top_2_contrib[2,6]
sec_R2 = one_day_top_2_contrib[3,6]

### 2021-04-15 submission/manuscript.tex
Add to results section

\textbf{5-day laxative treatment leads to prolonged \emph{C. difficile}
colonization in mice.} We compared PEG-treated mice to our standard 10
mg/kg clindamycin treatment, which temporarily renders the mice
susceptible to \emph{C. difficile}, with mice typically clearing
\emph{C. difficile} within 10 days post-infection (9, 19). All
PEG-treated mice were administered a 15\% PEG solution in the drinking
water for 5-days, one group was also treated with clindamycin, and one
group was allowed to recover for 10 days prior to challenge (Fig. 1A).
PEG treatment resulted in weight loss in all 3 groups of PEG-treated
mice, with the greatest change in weight observed on the fifth day of
PEG treatment (Fig. 1B). After either PEG, clindamycin, or PEG and
clindamycin treatment all mice were challenged with
10\textsuperscript{3} \emph{C. difficile} 630 spores. All treatments
rendered mice susceptible to \emph{C. difficile} colonization (Fig. 1C),
however PEG-treated mice remained colonized at a high level through 30
days post-infection. In contrast, the clindamycin-treated mice that
cleared \emph{C. difficile} within 10 days post-infection. Surprisingly,
mice were still susceptible to \emph{C. difficile} infection after
10-days of recovery from PEG treatment although \emph{C. difficile} was
not detectable in most of the group in the initial 5 days post-infection
(Fig. 1C). From 9 days post-infection onward, the median \emph{C.
difficile} stabilized for the 5-day PEG plus 10-day recovery group of
mice and remained high through 30 days post-infection (Fig. 1C). Thus,
osmotic laxative treatment alone was sufficient to render mice
susceptible to prolonged \emph{C. difficile} colonization and
PEG-treated mice remained susceptible for up to 10 days post PEG
treatment.

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since laxatives and
clindamycin have previously been shown to disrupt the murine microbiota
(ref), we hypothesized the different \emph{C. difficile} colonization
dynamics between mice treated with laxatives versus clindamycin were due
to their differential effects on the microbiota. We profiled the stool
microbiota over time by sequencing the V4 region of the 16S rRNA gene to
compare changes across treatment groups. We found time and treatment
group explained half of the observed variation between fecal communities
(PERMANOVA combined R\textsuperscript{2} = 0.50, \emph{P} \textless{}
0.001) with the remaining variation explained by interactions between
treatment group and other experimental variables including time, cage
effects, and sequencing preparation (Data Set S1, Sheet X). The impact
of cage effects has been well documented due to copraphagy and is
expected.

@@

\textbf{Figure 1. 5-day PEG treatment prolongs susceptibility and mice
become persistently colonized with \emph{C. difficile}.} A. Setup of the
experimental timeline for experiments with 5-day PEG treated mice.
Clindamycin was administered at 10 mg/kg by intraperitoneal injection.
15\% PEG 3350 was administered in the drinking water for five days. B.
Weight change from baseline weight in groups after treatment with PEG
and/or clindamycin, followed by \emph{C. difficile} challenge. C.
\emph{C. difficile} CFU/gram stool measured over time (N = 16-59 mice
per timepoint) via serial dilutions. The black line represents the limit
of detection for the first serial dilution. CFU quantification data was
not available for each mouse due to stool sampling difficulties
(particularly the day the mice came off of the PEG treatment) or early
deaths. For B-C, lines represent the median for each treatment group and
circles represent samples from individual mice. Asterisks indicate
timepoints where the weight change or CFU/g was significantly different
between groups by the Kruskal-Wallis test with Benjamini-Hochberg
correction for testing multiple timepoints. \newpage

>>>>

\textbf{5-day laxative treatment leads to prolonged \emph{C. difficile}
colonization in mice.} We compared mice treated with the osmotic
laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which
temporarily renders the mice susceptible to \emph{C. difficile}, with
mice typically clearing \emph{C. difficile} within 10 days
post-infection (9, 19). All PEG-treated mice were administered a 15\%
PEG solution in the drinking water for 5-days, one group was also
treated with clindamycin, and one group was allowed to recover for 10
days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss
in all 3 groups of PEG-treated mice, with the greatest change in weight
observed on the fifth day of PEG treatment and the mice regained most of
the weight five days after treatment (Fig. 1B). After either PEG,
clindamycin, or PEG and clindamycin treatment all mice were challenged
with 10\textsuperscript{3} \emph{C. difficile} 630 spores. All
treatments rendered mice susceptible to \emph{C. difficile} colonization
(Fig. 1C), however PEG-treated mice remained colonized at a high level
through thirty days post-infection. In contrast, the clindamycin-treated
mice that cleared \emph{C. difficile} within ten days post-infection.
Surprisingly, PEG-treated mice were still susceptible to \emph{C.
difficile} infection after 10-days of recovery from treatment although
\emph{C. difficile} was not detectable in most of the group in the
initial five days post-infection (Fig. 1C). From 9 days post-infection
onward, the median \emph{C. difficile} stabilized for the 5-day PEG plus
10-day recovery group of mice and remained high through 30 days
post-infection (Fig. 1C). Thus, osmotic laxative treatment alone was
sufficient to render mice susceptible to prolonged \emph{C. difficile}
colonization and PEG-treated mice remained susceptible for up to ten
days post-treatment.

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since laxatives and
clindamycin have previously been shown to disrupt the murine microbiota
(ref), we hypothesized the different \emph{C. difficile} colonization
dynamics between mice treated with laxatives versus clindamycin were due
to the two drugs having differential effects on the microbiota. We
profiled the stool microbiota over time by sequencing the V4 region of
the 16S rRNA gene to compare changes across treatment groups. We found
time and treatment group explained half of the observed variation
between fecal communities (PERMANOVA combined R\textsuperscript{2} =
0.50, \emph{P} \textless{} 0.001) with the remaining variation explained
by interactions between treatment group and other experimental variables
including time, cage effects, and sequencing preparation (Data Set S1,
Sheet X). Cage effects refer to the well-documented phenomenon that mice
housed in the same cages have similar microbial communities due to
copraphagy, we tried to minimize the impact of cage effects on our
experiment by breaking up cagemates when assigning mice to treatment
groups and primarily housing only two mice per cage. Importantly,
although we conducted a total of 5 separate experiments, the experiment
number and it's interaction with treatment group was not one of the
variables that significantly explained the observed variation in fecal
communities (Data Set S1, Sheet X). When we examine alpha diversity by
looking at Shannon diversity index over time, we see that although both
clindamycin and PEG treatments decrease diversity, the diversity index
is less in the groups of mice that received PEG treatment compared to
those that received clindamycin through thirty days post-infection.

@@

\textbf{Figure 1. 5-day PEG treatment prolongs susceptibility and mice
become persistently colonized with \emph{C. difficile}.} A. Setup of the
experimental timeline for experiments with 5-day PEG treated mice.
Clindamycin was administered at 10 mg/kg by intraperitoneal injection.
15\% PEG 3350 was administered in the drinking water for five days. B.
Weight change from baseline weight in groups after treatment with PEG
and/or clindamycin, followed by \emph{C. difficile} challenge. C.
\emph{C. difficile} CFU/gram stool measured over time (N = 16-59 mice
per timepoint) via serial dilutions. The black line represents the limit
of detection for the first serial dilution. CFU quantification data was
not available for each mouse due to stool sampling difficulties
(particularly the day the mice came off of the PEG treatment) or early
deaths. For B-C, lines represent the median for each treatment group and
circles represent samples from individual mice. Asterisks indicate
timepoints where the weight change or CFU/g was significantly different
between groups by the Kruskal-Wallis test with Benjamini-Hochberg
correction for testing multiple timepoints. The data presented are from
a total of 5 separate experiments. \newpage
